---
title: "Analyze data before correction"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=FALSE, warning=FALSE)
require("knitr")
```

```{r packages, include=FALSE}
# Required libraries
library(tidyr)
library(pheatmap)
library(factoextra)
library(plotly)
library(ggplot2)
```

### Load data and prepared fpkm values

```{r}
datasets = as.data.frame(scan("Stanford_datasets.txt",list(setname="",seqBatch="",species="",tissue=""),sep="\t"))
fpkmMat <- as.matrix(read.table('Stanford_datasets_fpkmMat.txt',header=FALSE,sep='\t'))
```

### Log transform data to normalize

```{r}
logTransformed_fpkmMat = log2(fpkmMat+1)
colnames(logTransformed_fpkmMat) <- datasets$setname
```

### Plot correlation heatmaps using different methods

```{r, echo=FALSE}
pheatmap(cor(logTransformed_fpkmMat)) 
pheatmap(cor(logTransformed_fpkmMat),clustering_method="average") 
pheatmap(cor(logTransformed_fpkmMat),clustering_method="single")
```

### Apply PCA method to uncorrected data

```{r}
transposeLogTransformed_fpkmMat = t(logTransformed_fpkmMat)
pca_proc <- prcomp(transposeLogTransformed_fpkmMat[,apply(transposeLogTransformed_fpkmMat, 2, var, na.rm=TRUE) != 0],scale=TRUE,center=TRUE)
```

### Check PCA statistics

```{r}
summary(pca_proc)
```

### Transfer PCA data to plots

```{r}
plotData = datasets[,c("setname","species","tissue")]
plotData$PC1 <- pca_proc$x[,1]
plotData$PC2 <- pca_proc$x[,2]
plotData$PC3 <- pca_proc$x[,3]
```

### Plot the first and the second principal components 

```{r, echo=FALSE}
p_scores <- ggplot(plotData, aes(x = PC1, y = PC2, shape=tissue))+geom_point(aes(color=species))+scale_shape_manual(values=c("adipose"=0,"adrenal"=1,"brain"=2,"sigmoid"=3,"heart"=4,"kidney"=5,"liver"=6,"lung"=7,"ovary"=8,"pancreas"=9,"small_bowel"=10,"spleen"=11,"testis"=12))
p_scores
```

### Plot the first and the second principal components with centroids

```{r}
plotData_pca <- prcomp(pca_proc$x[, -1])
fviz_pca_ind(plotData_pca,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = plotData$species, # color by groups
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Species",
             labs = "PCA before correction"
             )
```

### Plot the first, the second and the third principal components

```{r, echo=FALSE}
fig <- plot_ly(data = plotData, x = ~PC1, y = ~PC2, z = ~PC3, color=~species,shape=~tissue)
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                     yaxis = list(title = 'PC2'),
                     zaxis = list(title = 'PC3')))
fig
```

### Test for significance of correlations between the matched tissues PC values of human and mouse

```{r}
cor.test(plotData$PC1[1:13],plotData$PC1[14:26],method="pearson") 
cor.test(plotData$PC2[1:13],plotData$PC2[14:26],method="pearson") 
cor.test(plotData$PC3[1:13],plotData$PC3[14:26],method="pearson") 
```