---
title: "Analyze data after Combat correction"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=FALSE, warning=FALSE)
require("knitr")
```

```{r packages, include=FALSE}
# Required libraries
library(EDASeq)
library(edgeR)
library(sva)
library(ggplot2)
library(pheatmap)
library(plotly)
library(factoextra)
```

### Load data, prepared fpkm values and ortholog annonation

```{r}
datasets = as.data.frame(scan("Stanford_datasets.txt",list(setname="",seqBatch="",species="",tissue=""),sep="\t"))
rawCounts <- as.matrix(read.table('Stanford_datasets_rawCountsMat.txt',header=FALSE,sep='\t'))
geneDetails <- as.data.frame(scan("ortholog_GC_table.txt",skip=1,list(mouse_name="",mouse_GC = 0.0,human_name = "",human_GC=0.0)))
```

### Set columns names

```{r}
colnames(rawCounts) <- datasets$setname
rownames(rawCounts) <- geneDetails$human_name 
rownames(datasets) <- datasets$setname
rownames(geneDetails) <- geneDetails$human_name 
```

### Filter out lower 30% and mitochondrial genes

```{r}
rowSums = apply(rawCounts,1,function(x) sum(x))
quantile(rowSums,probs = 0.3) # result is 2947.9 
filter <- apply(rawCounts,1,function(x) sum(x)>2947.9 )
mt <- grep("mt-",geneDetails$mouse_name)
filteredNames <- setdiff(rownames(rawCounts[filter,]),rownames(rawCounts[mt,])) 
filteredRawCounts <- rawCounts[filteredNames,]
```

### Normalize data accounting for GC content

```{r}
GCnormCounts <- filteredRawCounts
GCnormCounts[,1:13] <- withinLaneNormalization(filteredRawCounts[,1:13],geneDetails[filteredNames,"human_GC"],which="loess",round=TRUE)
GCnormCounts[,14:26] <- withinLaneNormalization(filteredRawCounts[,14:26],geneDetails[filteredNames,"mouse_GC"],which="loess",round=TRUE)
```

### Depth normalize using TMM scaling factors

```{r}
origColSums <- apply(rawCounts,2,function(x) sum(x))
normFactors <- calcNormFactors(GCnormCounts,method='TMM')
colSums = apply(GCnormCounts,2,function(x) sum(x))
normalizedColSums <- origColSums
i <- 1
while (i<length(colSums)){
  normalizedColSums[i] <- origColSums[i]* normFactors[i]
  i <- i+1
}
meanDepth <- mean(normalizedColSums)
filteredDepthNormCounts <- GCnormCounts
i <- 1
while (i<ncol(filteredDepthNormCounts)){
 filteredDepthNormCounts[,i] <- (GCnormCounts[,i]/normalizedColSums[i])*meanDepth
 i <- i+1
 }
```

### Normalize using log transforming 

```{r}
logTransformedDepthNormCounts <- log2(filteredDepthNormCounts+1)
```

### Apply Combat method to correct batch effect

```{r}
meta <- data.frame(seqBatch = datasets$seqBatch,tissue=datasets$tissue,species=datasets$species)
design <- model.matrix(~1,data=meta)
combat <- ComBat(dat= logTransformedDepthNormCounts,batch=meta$seqBatch,mod=design,par.prior=TRUE)
```

### Plot correlation heatmaps using different methods

```{r, echo=FALSE}
pheatmap(cor(combat)) 
pheatmap(cor(combat),clustering_method="average")
pheatmap(cor(combat),clustering_method="single") 
```

### Apply PCA method to corrected data

```{r}
transposed_combat <- t(combat)
pca_proc <- prcomp(transposed_combat[,apply(transposed_combat, 2, var, na.rm=TRUE) != 0],scale=TRUE,center=TRUE,retX=TRUE)

```

### Check PCA statistics

```{r}
summary(pca_proc)
```

### Transfer PCA data to plots

```{r}
plotData = datasets[,c("setname","species","tissue")]
plotData$PC1 <- pca_proc$x[,1]
plotData$PC2 <- pca_proc$x[,2]
plotData$PC3 <- pca_proc$x[,3]
```

### Plot the first and the second principal components 

```{r, echo=FALSE}
p_scores <- ggplot(plotData, aes(x = PC1, y = PC2, shape=tissue))+geom_point(aes(color=species))+scale_shape_manual(values=c("adipose"=0,"adrenal"=1,"brain"=2,"sigmoid"=3,"heart"=4,"kidney"=5,"liver"=6,"lung"=7,"ovary"=8,"pancreas"=9,"small_bowel"=10,"spleen"=11,"testis"=12))
p_scores
```

### Plot the first and the second principal components with centroids

```{r, echo=FALSE}
plotData_pca <- prcomp(pca_proc$x[, -1])
fviz_pca_ind(plotData_pca,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = plotData$species, # color by groups
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Species"
             )
```

### Plot the first, the second and the third principal components

```{r, echo=FALSE}
fig <- plot_ly(plotData, x = ~PC1, y = ~PC2, z = ~PC3, color=~species,shape=~tissue)
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(xaxis = list(title = 'PC1'),
                     yaxis = list(title = 'PC2'),
                     zaxis = list(title = 'PC3')))
fig
```

### Test for significance of correlations between the matched tissues PC values of human and mouse

```{r}
cor.test(plotData$PC1[1:13],plotData$PC1[14:26],method="pearson") 
cor.test(plotData$PC2[1:13],plotData$PC2[14:26],method="pearson")
cor.test(plotData$PC3[1:13],plotData$PC3[14:26],method="pearson") 
```
